Environmental suitability throughout the late Quaternary explains population genetic diversity.

### 2023-08-08 manuscript_english.Rmd

---
title: Suitability prevalence area index in late quaternary explains genetic diversity in Tassel eared Squirrels
author: "Norma Hernandez, Angel Robles Nathan Upham"
date: "2023-06-27"
includes:
  in_header: preamble.tex
output: 
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex 
    keep_tex: true
bibliography: R21_citation.bib
---

@@

Genetic variation among populations exhibits biogeographic patterns. In this paper we establish a theoretical framework based on environmental Grinellian niches as a tool for modeling species distributions, as well as their projected past environmental scenarios to relate the present evolutionary history in both the geographic distribution and genetic structure of their populations. Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework with the help of the joint analysis of spatial, environmental and genetic data, with tools that are widely distributed among biogeographers and evolutionary biologists. Our results reveal that when we have case of isolation by distance, suitability prevalence along time corresponds to the fixation index of populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower $F_{st}$ value. This framework allows us to add a biogeographic explanation to the results obtained with population genetic methods and to generate maps of this structure as tools to support conservation with a perspective that integrates both population genetics and historical patterns of species distribution.

@@

The current distributions of species do not always correspond to their historical distributions over evolutionarily significant periods. This is because environmental conditions are not static over time; species tend to distribute where conditions are most favorable.

It follows from this relationship between environment and species distribution that population size also varies with time. This variation in population size is related to the effective population size $N_e$ [@karlin1968rates], so that an index reflecting changes in environmental conditions in geography might be expected to be related to $N_e$ and thus to indicators of population structure such as the fixation index $F_{st}$. Thus, it is possible to relate patterns of changes in the distribution of species to the genetic structure of their populations using a statistical model that explains this relationship.

With this approach we can predict the geographic pattern of population structure from environmental information. This approach is strongly driven by advances in currently available climate simulations [@leonardi2023pastclim; @krapp2021statistics], as well as next generation sequencing data, and supported by both ecological niche [@thorup2021response; @nogues2009predicting] and population genetics theories [@lira2014relationship].

In this work we propose a method to find the Suitability Prevalence Area (SPA) as an index with a double purpose: 1) to find endemic areas to delimit the historical distribution of the species and 2) to explain the patterns of genetic diversity.

To obtain the SPA, we performed a historical reconstruction of the geographical range back to 120 000 BC at 2 000 year intervals and recorded the environmental suitability at each site.

Subsequently, we delimited historical endemic areas to locations where the prevalence of suitability remained at 90% during this period. From the fixation index calculated with respect to populations in the historical endemic areas, a statistical model of the fixation index as a function of SPA was performed. With this statistical model, the fixation index values were projected to the current distribution to obtain a map with the geographic pattern of this index.

@@

We define $\vec{e}_i =(e_1, e_2, ..., e_i )$ as a vector of environmental variables, that is, a point in the space of $v$ environmental variables. Then $R_i(\vec{e}_i)$ is a function that relates evolutionary fitness to environment,
that is, for a species with non-overlapping generations, the population's net reproductive rate $R_i$ is a measure of evolutionary fitness that depends on environmental conditions.

That is, the sites in the geography $\mathbf{G}_o$ that present positive growth rates are related to favorable conditions.
If we take the set of places in ambient space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.

This is a way of expressing the fundamental niche as a function of environmental conditions. So we can map each point of geography to ambient space. Thus, for each site in the geography there will correspond a site in the environmental space with different environmental conditions and different growth rates. However all sites within the region defined by $\mathbf{N}_F$ will have viable populations and positive rates. However, we assume that the environment is dynamic and each geographic point is static, so for each site in geographic space this growth rate depends on the environment and therefore is different at each time.

Thus locally the growth rate will be related to the local population size as long as it is within the boundary defined by $\mathbf{N}_F$. Therefore, the local population size as a function of the intrinsic rate that depends on the change in environmental conditions over time is:
$N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)$. That is, the locally effective population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of $R_0$ as a set of trajectories for each point from geographic space to environmental space. Considering this rate fixed $\mathbf{R}_0$ when the population grows, that is, $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > 1 \}$ and therefore any local population will have a positive growth rate.

@@

This equation was proposed by Soberon and Peterson @soberon2020shape to relate demographics to a function of evolutionary fitness and environment. In that study they tested the validity with monthly variables and empirical values on a short time scale. relating life history parameters and the environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we relate this growth rate to an environmental suitability function bounded between $0$ and $1$, for example, for a unimodal multivariate pdf, it would be equivalent to the distance from the maximum of the function. That is, the fitness-related environmental suitability can be any function such that it is $0$ if $R_i(\vec{e}_i) > 1$ and $1$ when $R$ is evaluated at the expected value of $\vec {e}$ for all values of $R_i(\vec{e}_i) > 1$. In a practical way, $\mathbf{N}_F$ is approximated from points of presence in the geography obtained from databases, numerically approximating a convex surface with a surface $\mathbf{N}_R$ of the environmental space.

@@

The goal of this theoretical approach is not to know the shape of the function of $R_0$ but rather to find this local variation relating this value of $R_0$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, @phillips2006maximum), i.e. :

$$
N_i \propto  S_i(\vec{e}_i)
$$
Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$

@@

On the other hand, loss of heterozygosity may be related to the amount of genetic variation present in the absence of mutation and selection, so theoretically, one would expect a strong correlation between effective population size and heterozygosity on the basis of population genetics. We start from the assumption that in the absence of gene flow between subpopulations, the rate of fixation by isolation depends on the effective population size $N_e$ when subpopulations diverge from a common ancestor in $t$ generations. [@hedrick2005standardized: @wright1943isolation]:

@@

Moreover, the effective population can be estimated as the harmonic mean of the population in a demographic time series [@karlin1968rates] for $\tau$ generations $\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)$. We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close.

@@

$$
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
$$
If we consider all sites in the available geography, we call this harmonic mean *Suitability Prevalence Area* $\mathbf{S} = \{S_i | S_i > k\}$ (SPA) since we expect that areas where the $F_{st}$ is lower are in areas where the prevalence over time is higher, i.e. environmental conditions remain suitable for a significant period of time and therefore no effects of genetic drift of populations have occurred. Conversely, it is expected that in populations with a high fixation index value, suitability will not have prevailed over the same time interval even though conditions currently exist and populations are present.

@@

1. model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
3. Characterize past analogous climates and retain projections only for sites with climates analogous to the present.
4. Generate a harmonic mean suitability map for all time slices.

@@

- Obtain georeferenced points of genetically structured populations with fixation index values.
- Extract values from the SPA map with these points. 
- A linear model of the fixation index value with respect to the SPA is performed.
- With the linear model all SPA values of suitability are interpolated.
- A map is generated with all the values interpolated by the model.

@@

@@

#### Analogy of environmental conditions 
To be certain that areas in the past are climatically analogous to the present we implemented the mobility-orient parity (MOP) test. This method generates a value from $0$ to $1$ according to the environmental analogy of past conditions to the present to eliminate geographic areas in the projections to the past outside of climates analogous to the present [@owens2013constraints].

@@

### Fixation index as a functino of SPA

@@

With the current availability of occurrence data and models of potential distribution projected to past climate scenarios it is possible to find geographic patterns of differentiation in populations at an acceptable resolution (half a degree) that allows us to define sites where the species can potentially be found but share different degrees of genetic diversity.

Projections of genetic diversity in geography have been made in previous studies @van2012mapping. However in a general way they are extrapolations of observations, without a predictive model, which with some explanatory variable behind can explain in a statistically convincing way the information obtained from population genetics.

@@

Thanks to the theoretical support that we have briefly outlined in this paper, it is possible to explain the explanatory value of the Coefficient of determination for our linear model ($R^2 = 0.79 $), which is information coming from two completely different sources, without any correlation, but always guided by the premise of the evolutionary process as causal.

@@

The generalization of our protocol and its implementation in tools widely used by biogeography allow its direct implementation in many case studies with available information, as well as the design of new questions, e.g. species assemblages under prevalence of environmental suitability. A research path was also established on whether the process of environmental stability over time is related to biodiversity.

>>>>

---
title: Environmental suitability throughout the late Quaternary explains population genetic diversity. 
author: "Norma Hernandez, Angel Robles Nathan Upham"
date: "2023-06-27"
includes:
  in_header: preamble.tex
output: 
  pdf_document:
    fig_caption: yes
    latex_engine: xelatex 
    keep_tex: true
bibliography: R21_citation.bib
---

@@

Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established. In this paper, we establish a theoretical framework based on projecting environmental Grinellian niches back through time to relate the present geographic distribution of population genetic structure to a given species' historical evolutionary context.

Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework directly. We use as a case study the Tassel-eared Squirrel (*Sciurus aberti*) to jointly analyze spatial, environmental, and genetic data to predict the historical endemic area of this species. Our results reveal that in cases of genetic isolation by geographic distance, the prevalence of environmental suitability over time corresponds to the genetic fixation index of populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower$F_{st}$ value.

This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a biogeographic explanation to the results obtained with population genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to support biodiversity conservation efforts. Overall, this work advances a perspective that integrates population genetics with historical patterns of species distribution.

@@

Current species distributions do not always correspond to their
historical distributions over evolutionarily significant periods. Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., @lessa2003genetic). This is because environmental conditions are highly dynamic over time, and species
tend to track those environments, distributing where conditions are most favorable. That is, species environmental niches tend to be conserved (@peterson2011ecological).

It follows from this relationship between environment and species distribution that population size also varies with time. This variation in population size is related to the effective population size $N_e$
[@karlin1968rates], so that an index reflecting changes in environmental
conditions in geography might be expected to be related to $N_e$ and thus to indicators of population structure such as the fixation index
$F_{st}$. It should therefore be possible to relate patterns of change in the distribution of species to the genetic structure of their populations using a statistical model that explains this relationship.

With this approach we expect to predict the geographic pattern of population-genetic structure from abiotic environmental information alone. This approach is strongly driven by advances in currently available climate simulations [@leonardi2023pastclim; @krapp2021statistics], as well as next-generation sequencing data, and supported by both ecological niche [@thorup2021response; @nogues2009predicting; @lira2014relationship] and population genetics theories [@lynchi2018]. Importantly, these predicted patterns of population structure will serve as a type of \textit{a priori} hypothesis for the influence of abiotic factors alone, with the empirical detection of deviations from this hypothesis indicating that biotic factors such as competition or predation from other species have also impacted the observed population structure.

In this work, we propose a method to find an index that we call the Suitability Prevalence Area (SPA) that is the area of highest environmental suitability for a given species over a historical time period. The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) to explain the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.

As a case study, we consider Tassel-eared Squirrels (\textit{Sciurus aberti}), a species of rodent that is currently distributed in disjunct patches from the southern Rocky Mountains in the United States to the northern Sierra Madre Occidental in Mexico for which it is possible to find reliable information on both the genetic structure of its populations and its current distribution [@bono2018genome; @burgin2018many]. To obtain the SPA for \textit{S. aberti}, we overlapped historical reconstructions of the species' geographical distribution estimated every 2,000 years from the present day until 120,000 years ago, recording the environmental suitability in each time interval at each modeled site.

Subsequently, we calculated the fixation index ($F_{st}$) of the genetic relatedness among populations and compared that to the SPA index of historical endemism, constructing a statistical model of the fixation index as a function of the SPA. With this explanatory model, the $F_{st}$ values were projected to the current geographic distribution of \textit{S. aberti} to obtain a predictive map of expected $F_{st}$ values over space.

@@

We define \(\vec{e}_i =(e_1, e_2, ..., e_i )\) as a vector of
environmental variables, that is, a point in the space of \(v\)
environmental variables. Then \(R_i(\vec{e}_i)\) is a function that
relates evolutionary fitness to environment. That is, for a species with
non-overlapping generations, the population's net reproductive rate 
\(R_i\) is a measure of evolutionary fitness that depends on
environmental conditions.

In other words, $\mathbf{G}_o$ is the set of sites in the geography that present positive growth rates related to favorable conditions. If we take the set of places in environmental space where these values are greater than a threshold, we can define a volume in multidimensional space such that $R(\vec{e})$ is always positive and greater than this threshold, that is $\mathbf{N}_F(t) = \{\vec{e} | R_0(\vec{e}(t)) > k \}$.

This is a way of expressing the fundamental niche \(\mathbf{N}_F\) as a function of environmental conditions. Thus, for each site in the geography there will
correspond a site in the environmental space with different
environmental conditions and different growth rates. However, all sites
within the region defined by \(\mathbf{N}_F\) will have viable
populations and positive growth rates. Nonetheless, we assume that the environment is dynamic and each geographic point is static, so for each site in
geographic space this growth rate depends on the environmental conditions and
therefore growth rate changes in value at each time interval.

Locally, the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic growth rate, which depends on the change in environmental conditions over time, can be expressed as:
\(N_i(t+1) = R(\vec{e_i}(t) ) N_i(t)\). That is, the locally population size is a function of ideal environmental conditions.

We start by considering this defined volume of the set of \(R_0\) as a
set of trajectories for each point from geographic space to
environmental space. Considering this rate fixed \(\mathbf{R}_0\) when
the population grows, that is,
\(\mathbf{N}_F(t) = \{\vec{e} | R(\vec{e}(t)) > 1 \}\) and therefore
any local population will have a positive growth rate.

@@

This equation was proposed by Soberon and Peterson [-@soberon2020shape] to relate demographics to a function of evolutionary fitness and environment. In that study they tested the validity with monthly variables and empirical values on a short time scale. relating life history parameters and the environmental suitability of niche models.

This set can be seen as a convex volume in environmental space and we
relate this growth rate to an environmental suitability function bounded
between \(0\) and \(1\), for example, for a unimodal multivariate probability density function,
it would be equivalent to the distance from the maximum of the function.
That is, the fitness-related environmental suitability can be any
function such that it is \(0\) if \(R_i(\vec{e}_i) > 1\) and \(1\) when
\(R\) is evaluated at the expected value of \(\vec {e}\) for all values
of \(R_i(\vec{e}_i) > 1\). In a practical way, \(\mathbf{N}_F\) is
approximated from points of presence in the geography obtained from
databases, numerically approximating a convex surface with a surface
\(\mathbf{N}_R\) of the environmental space.

@@

The goal of this theoretical approach is not to know the shape of the function of $R$ but rather to find this local variation relating this value of $R$ and the function $\mathbf{N}_F$ with a function $ S_i(\ vec{e}_i)$ of environmental suitability that can be estimated from known methods of applied mathematics and machine learning (for example the case of Maxent, [@phillips2006maximum]), i.e. :

$$
N_i \propto  S_i(\vec{e}_i)
$$
Population size locally is a function of environmental suitability, where if suitability is $0$ the population is $0$ and if suitability is $1$ the population has a maximum growth rate $R_i(\vec{e}_i)$ in some point of the environmental space.

@@

On the other hand, loss of heterozygosity may be related to the amount of genetic variation present in the absence of mutation and selection, so theoretically, one would expect a strong correlation between effective population size and heterozygosity on the basis of population genetics. We start from the assumption that in the absence of gene flow between subpopulations, the rate of fixation by isolation depends on the effective population size $N_e$ when subpopulations diverge from a common ancestor in $t$ generations. [@hedrick2005standardized: @wright1943isolation, @lynchi2018]:

@@

Moreover, the effective population can be estimated as the harmonic mean of the population in a demographic time series [@karlin1968rates] for $\tau$ generations $\frac{1}{N_e} \approx \frac{1}{\tau}\Big({\frac{1}{N_1} + \frac{1}{N_2} + ... + \frac{1}{N_\tau} } \Big)$. We substitute the harmonic mean into the above equation, considering that although the number of generations is not exactly equal to the time of the fixation index, they are expected to be very close:

@@

$$
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
$$
If we consider all sites in the available geography, we call this
harmonic mean \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) (SPA) since we expect that areas
where the \(F_{st}\) is lower are in areas where the prevalence over
time is higher, i.e.~environmental conditions remain suitable for a
significant period of time and therefore no effects of genetic drift of
populations have occurred. Conversely, it is expected that in
populations with a high fixation index value, suitability will not have
prevailed over the same time interval even though conditions currently
exist and populations are present.

@@

1. Model the potential distribution of a taxon with present-day environmental conditions and obtain an environmental suitability map.
2. Project the potential distribution model to various environmental scenarios in the past with constant time steps and obtain the environmental suitability for each time scenario.
3. Characterize past analogous climates and retain projections only for sites with climates analogous to the present.
4. Generate a harmonic mean suitability map for all time slices.

@@

- Obtain georeferenced points of genetically structured populations with fixation index values.
- Extract values from the SPA map with these points. 
- Perform a linear model of the fixation index value with respect to the SPA.
- With the linear model all SPA values of suitability are interpolated.
- Generate a map with all the values interpolated by the model.

@@

As a case study, we consider squirrels (\emph{Sciurus aberti}), a species
currently distributed in disjunct patches from the southern Rocky
Mountains in the United States to the northern Sierra Madre Occidental
in Mexico for which it is possible to find reliable information on both
the genetic structure of its populations and its current distribution
[@bono2018genome; @burgin2018many]. In order to select this species,
we list the following considerations:

@@

@@

### Fixation index as a function of SPA

@@

With the current availability of occurrence data and models of potential
distribution projected to past climate scenarios it is possible to find
geographic patterns of differentiation in populations at an acceptable
resolution (half a degree) that allows us to define sites where the
species can potentially be found but share different degrees of genetic
diversity..

Projections of genetic diversity in geography have been made in previous
studies [@van2012mapping]. However in a general way they are
extrapolations of observations, without a predictive model, which with
some explanatory variable behind can explain in a statistically
convincing way the information obtained from population genetics.

@@

Thanks to the theoretical support that we have briefly outlined in this
paper, it is possible to explain the explanatory value of the
coefficient of determination for our linear model, $R^{2} = 0.79$,
which is information coming from two completely different sources, but always guided by the premise of the evolutionary process as causal.

@@

The generalization of our protocol and its implementation in tools
widely used by biogeography allow its direct implementation in many case
studies with available information, as well as the design of new
questions, e.g.~species assemblages under prevalence of environmental
suitability. A research path was also established on whether the process
of environmental stability over time is related to biodiversity.

It is also noteworthy to mention that in previous taxonomic descriptions, it was taken into account that the distribution of Sciurus aberti was only related to the distribution of Arizona pinus ponderosa. However, there is a consensus on the range of distribution of this pine, which is limited to the southern United States [@hess2020mexican]. Therefore the known distribution of Sciurus aberti is in mixed coniferous forests. This plasticity in the squirrel-pine interaction explains the expansion of the species in interglacial periods and its distribution in warmer periods, such as the current one. However, the low genetic diversity found in places where the squirrel species interact with few pine species could put it at greater risk of climate change that constricts the distribution of p. p. Arizona, especially in the Sky Island region of Arizona.
